\documentclass[english,phd]{diploma}

% Packages
% enhanced support for quotation marks, support for biblatex package
\usepackage[autostyle=true]{csquotes}
\usepackage[
    backend=biber,
    giveninits=true, % abbreviate author's first names in bibliography
    style=numeric,
    alldates=iso,
    maxnames=999,    % for my own publications
    maxbibnames=3,   % for bibliography
    defernumbers=true,
    urldate=short,
    sorting=none,
    seconds=true
]{biblatex}
\usepackage{dcolumn} % numeric column type
%\usepackage{subfig} % subtables and subfigures

\usepackage{subcaption} % subfigure
\captionsetup{compatibility=false}

\newcolumntype{T}{>{\footnotesize}p{6cm}} % Estee table

\usepackage[cpp]{diplomalst}
\usepackage{hyperref}
%\usepackage{todonotes}
\usepackage{amsmath}
% \mathlarger
\usepackage{relsize}
% \mathbb
\usepackage{amssymb}
\usepackage{tikz}
\usetikzlibrary{graphs}
\usetikzlibrary{positioning}
\usetikzlibrary{arrows.meta}
\usetikzlibrary{automata}
\usetikzlibrary{math}
% Source code highlighting
\usepackage[outputdir=../out,newfloat]{minted}
% Multiple footnotes at one place
\usepackage[multiple]{footmisc}
% Autoref with capitalization (need to use Cref)
\usepackage[nameinlink]{cleveref}
% Horizontal margin configuration in description/enumerate/itemize
\usepackage{enumitem}
% SI units (\micro\second, \mebi\byte)
\usepackage{siunitx}
% HQ comparison table
\usepackage{array}
\usepackage{fullpage}
\usepackage{makecell}
\usepackage{fontawesome}
\usepackage{threeparttable}
\usepackage{wasysym}

% Abbreviations (\gls, \glspl, \acrfull, \acrshort, \acrlong)
\usepackage[
    acronym,        % enable acronyms
    nomain,         % disable main glossary, we only use acronyms
    notranslate     % enable overriding \acronymname without involving babel
]{glossaries-extra}

%\glsdisablehyper    % disable hyperlinks
\renewcommand*{\acronymname}{List of Abbreviations}

\usemintedstyle{vs}
\setminted{tabsize=4}
% [escapeinside=||]
% [tabsize=4]
% [fontsize=\small]

% Fix arrow tips
\tikzset{tips=proper}

% Commands
\newcommand{\comment}[1]{}
\newcommand{\workshare}[1]{\emph{#1}}
\newcommand{\footnoteurl}[1]{\footnote{\url{#1}}}
\newcommand{\Autoref}[1]{\Cref{#1}}

% Tools
\newcommand{\estee}{\textsc{Estee}}
\newcommand{\rsds}{\textsc{RSDS}}
\newcommand{\hyperqueue}{\textsc{HyperQueue}}
\newcommand{\autoalloc}{\emph{autoalloc}}
\newcommand{\Autoalloc}{\emph{Autoalloc}}
\newcommand{\hq}{\textsc{HQ}}
\newcommand{\gnuparallel}{\textsc{GNU parallel}}
\newcommand{\hypershell}{\textsc{HyperShell}}
\newcommand{\dask}{\textsc{Dask}}
\newcommand{\ray}{\textsc{Ray}}
\newcommand{\parsl}{\textsc{Parsl}}
\newcommand{\pycompss}{\textsc{PyCOMPSs}}
\newcommand{\pegasus}{\textsc{Pegasus}}
\newcommand{\balsam}{\textsc{Balsam}}
\newcommand{\autosubmit}{\textsc{AutoSubmit}}
\newcommand{\fireworks}{\textsc{FireWorks}}
\newcommand{\merlin}{\textsc{Merlin}}
\newcommand{\snakemake}{\textsc{SnakeMake}}

% Definitions
\newcounter{taskdefinition}

% #1 = label
\newcommand{\makedef}[1]
{\refstepcounter{taskdefinition}%
[\textbf{Definition~\thetaskdefinition{}}]\label{#1}}
\newcommand{\defref}[1]{\hyperref[#1]{Definition~\ref*{#1}}}

% Specific definitions
\newcommand{\fntaskres}{Res_t}
\newcommand{\fntaskresfrac}{\fntaskres^{fr}}
\newcommand{\fntaskresvar}{\fntaskres^{var}}
\newcommand{\fnworkerres}{Res_w}
\newcommand{\fnworkerresident}{\fnworkerres^{id}}
\newcommand{\fntaskresassigned}{RA}
\newcommand{\fntaskvarassigned}{RV}
\newcommand{\fnworkertaskassigned}{WX}
\newcommand{\fntaskworkerassigned}{X}
\newcommand{\fntaskstart}{S}
\newcommand{\fntaskfinish}{F}
\newcommand{\fntaskresvarcount}{RC}
\newcommand{\scheduleworker}{S_w}
\newcommand{\scheduleorder}{S_o}
\newcommand{\setresourcekinds}{RK}
\newcommand{\setresourcekindstaskgraph}{RK_t}
\newcommand{\setresourcekindscluster}{RK_w}
\newcommand{\setresourcekindsident}{RK_w^{id}}
\newcommand{\setresourceelements}{RE}
\newcommand{\powerset}[1]{\mathcal{P}(#1)}
\newcommand{\resourcerequirement}[2]{$#1\colon T \times \setresourcekindstaskgraph \rightarrow \mathbb{#2}_{\geq{}0}$}
\newcommand{\timedomain}{\mathbb{R}_{\geq{}0}}
\newcommand{\taskgraphinner}[1]{(T, O, E, #1, \fntaskres)}
\newcommand{\taskgraphexecution}{$E = (G, C, \fntaskworkerassigned)$}

\graphicspath{{./imgs/}}

% TikZ styles
\tikzset{%
    task/.style={circle, draw},
    data/.style={rectangle, draw, rounded corners},
    arrow/.style={draw, thick, {->}}
}

\renewcommand\lstlistlistingname{List of Source Code Listings}

% Metadata
\ThesisAuthor{Ing. Jakub Beránek}

\ThesisSupervisor{Ing. Jan Martinovič, Ph.D.}

\CzechThesisTitle{Ergonomie a efektivita workflow na HPC klastrech}
\EnglishThesisTitle{Ergonomics and efficiency of workflows on HPC clusters}

\SubmissionYear{2024}

\CzechAbstract{Tato práce se zabývá spouštěním grafů úloh na vysoce výkonných systémech (superpočítačích),
se zaměřením na efektivní využití výpočetních zdrojů a poskytnutí ergonomických rozhraní pro návrh
a spouštění grafů úloh. Programování na základě úloh je oblíbeným způsobem pro definici vědeckých soustav
výpočtů, které jsou určeny pro spouštění na distribuovaných systémech. Nicméně spouštění těchto úloh na superpočítačích
přináší unikátní výzvy, například problémy s výkonem způsobené značným rozsahem úloh nebo problematickou interakci úloh
se systémy pro správu alokací na superpočítačích, jako jsou například PBS (Portable Batch System) nebo Slurm. Tato práce
zkoumá, jaké jsou hlavní problémy ovlivňující spouštění úloh v této oblasti a navrhuje různé přístupy,
které by měly pomoci tyto problémy částečně či zcela vyřešit, a to jak v oblasti výkonu, tak i ergonomie vývoje.

Tato práce poskytuje tři hlavní přínosy. Prvním z nich je prostředí pro simulaci spouštění grafů úloh,
které umožňuje jednoduché experimentování a měření různých plánovacích algoritmů. Toto prostředí bylo
použito pro provedení rozsáhlé studie kvality různých plánovačů úloh. Dále práce analyzuje
výkonnostní charakteristiku moderního nástroje pro spouštění úloh \dask{}, a poskytuje alternativní
implementaci \dask{} serveru, která výrazně zvyšuje jeho efektivitu v případech, které vyžadují vysoký výkon.
Hlavním přínosem práce je metoda pro plánování úloh a správu zdrojů, která umožňuje jednoduché spouštění grafů úloh na heterogenních superpočítačích, které zároveň maximalizuje využití dostupných výpočetních zdrojů. Práce také poskytuje referenční implementaci
využívající této metody v rámci nástroje \hyperqueue{}, který je dostupný jako software s otevřeným zdrojovým kódem pod licencí MIT (Massachusetts Institute of Technology) na adrese \url{https://github.com/it4innovations/hyperqueue}.}
\CzechKeywords{distribuované výpočty, výpočetní grafy, heterogenní zdroje, vysoce výkonné počítání}

\EnglishAbstract{This thesis deals with the execution of task graphs on High-performance Computing
(HPC) clusters (supercomputers), with a focus on efficient usage of hardware resources and ergonomic interfaces for
task graph submission. Task-based programming is a popular approach for defining scientific workflows that can be computed on distributed
clusters. However, executing task graphs on supercomputers introduces unique challenges, such as performance
issues caused by the large scale of HPC workflows or cumbersome interactions with HPC allocation
managers like PBS (Portable Batch System) or Slurm. This work examines what are the main challenges
in this area and how do they affect task graph execution, and it proposes various approaches for alleviating
these challenges, both in terms of efficiency and developer ergonomics.

This thesis provides three main contributions. Firstly, it provides a task graph simulation environment
that enables prototyping and benchmarking of various task scheduling algorithms, and performs a comprehensive
study of the performance of various task schedulers using this environment. Secondly, it analyzes the
bottlenecks and overall performance of a state-of-the-art task runtime \dask{} and provides an implementation
of an alternative \dask{} server which significantly improves its performance in HPC use-cases. And
primarily, it introduces a unified meta-scheduling and resource management design for effortless execution of task graphs on heterogeneous HPC clusters that facilitates efficient usage of hardware resources. It also provides a reference implementation of this design within an HPC-tailored task runtime called \hyperqueue{}, which is available as open-source software under the MIT (Massachusetts Institute of Technology) license at \url{https://github.com/it4innovations/hyperqueue}.}
\EnglishKeywords{distributed computing, task graphs, heterogeneous resources, hpc}

\Acknowledgement{I would like to thank my supervisor, Jan Martinovič, for his advice.
I would also like to thank Ada Böhm, Vojtěch Cima and Martin Šurkovský, who have co-authored several
publications with me and supported me during my PhD studies. I am especially grateful to Ada Böhm
for her mentorship and constant readiness to provide both research and technical guidance. I would like
to express my gratitude to all the wonderful people that I met during my internship at the SPCL lab at ETH Zurich.
Furthermore, my thanks also go to Vanessa DeRhen for proofreading this thesis.

\vspace{2mm}The development of \estee{} was supported by several projects.
It has received funding from the European High-Performance Computing Joint Undertaking (JU)
under grant agreement No 955648. This work was supported by the Ministry of Education, Youth and Sports of the Czech Republic the e-INFRA CZ (ID: 90140) and ACROSS (ID: MC2104) projects.

The development of \hyperqueue{} was supported by several projects.
It has received funding from the European High-Performance Computing Joint Undertaking (JU) under grant agreement No 956137.
This work was supported by the Ministry of Education, Youth and Sports of the Czech Republic through the e-INFRA CZ (ID: 90254) and LIGATE (ID: MC2102)
projects.

\vspace{2mm}Last but not least, I thank my wife Jana for her support and endless patience.}

\addbibresource{references.bib}
\addbibresource{publications.bib}

\makeglossaries
\setabbreviationstyle[acronym]{short-long}
\input{acronyms}

\begin{document}
\MakeTitlePages

\begin{spacing}{1}
	%\glsaddall     % Automatically add all defined acronyms
	%\addcontentsline{toc}{chapter}{List of Abbreviations}
	\printglossaries

	\listoffigures
	\clearpage

	\listoftables
	\clearpage

	\lstlistoflistings
	%\addcontentsline{toc}{chapter}{List of Source Code Listings}
	\clearpage
\end{spacing}

% Make sure that related publications are first in the bibliography list
\nocite{estee, rsds, hyperqueue, ligate}

\begin{spacing}{1.1}
%	\chapter{Introduction}
%	\label{ch:Introduction}
%	\input{chapters/introduction}
%
%	\chapter{Parallel and distributed computing}
%	\label{ch:distributed-computing}
%	\input{chapters/distributed-computing}

	\chapter{Task-based programming}
	\label{ch:taskgraphs}
	\input{chapters/taskgraphs}

%	\chapter{State of the Art}
%	\label{ch:sota}
%	\input{chapters/sota-challenges}
%
%	\chapter{Task scheduling analysis}
%	\label{ch:estee}
%	\input{chapters/estee}
%
%	\chapter{Task runtime optimization}
%	\label{ch:rsds}
%	\input{chapters/rsds}
%
	\chapter{Task graph meta-scheduling}
	\label{ch:hyperqueue}
	\input{chapters/hyperqueue}
%
%	\chapter{Conclusion}
%	\label{ch:conclusion}
%	\input{chapters/conclusion}
\end{spacing}

\begin{spacing}{1}
	\chapter*{List of own publication activities}
	\label{ch:listofstudentsownpublicationactivities}
	\addcontentsline{toc}{chapter}{\nameref{ch:listofstudentsownpublicationactivities}}

	\input{chapters/publications-generated}

	%	\chapter*{List of projects}
	%	\label{ch:listofprojects}~\addcontentsline{toc}{chapter}{\nameref{ch:listofprojects}}
	%	\input{chapters/projects}

	% Bibliography
	\renewcommand*{\bibfont}{\small}
	\printbibliography[heading=bibintoc, title={Bibliography}]
\end{spacing}

\appendix
\input{chapters/appendix}

\end{document}
